---
title: "Reproducible Research: Peer Assessment 1"
author: "N Touheed"
date:   "7/13/2020"
output: 
  html_document:
    keep_md: true
---
<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## R Markdown

## Peer-graded Assignment: Course Project 1

### Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


Now let us go to load and preprocess the data. And show the code that is needed to

1. Load the data (i.e. read.csv()) 
2. Process/transform the data (if necessary) into a format suitable for our analysis
And find the mean total number of steps taken per day.

## Loading and preprocessing the data
```{r}
# We first unzip the zipfile

unzip(zipfile="repdata_data_activity.zip")

# Now bring the data into a data frame called activity 

activity <- read.csv("activity.csv")

# removing NA's from activity

activity <- activity[ with (activity, {!(is.na(steps))}),]

summary(activity)
```

## What is mean total number of steps taken per day?

For this part of the assignment, we have ignored the missing values in the dataset. And would like to perform the following activities:

1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r `echo = FALSE`}
library(dplyr)

eachDay <- group_by(activity, date)
stepsTakenEachDay <- summarise(eachDay, total = sum(steps))
head(stepsTakenEachDay)

hist(stepsTakenEachDay$total, main="Histogram of the total number of steps taken each day", xlab="Total number of steps taken each day")

cat("Mean value of total number of steps each day is", mean(stepsTakenEachDay$total))
cat("Median value of total number of steps each day is", median(stepsTakenEachDay$total))

```

## What is the average daily activity pattern?

We next like to addrees the following question:

What is the average daily activity pattern?

To answer this we do the following:

1.	Make a time series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2.	To find the 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps

```{r}
# preparing data for time series plot

intervalSteps <- aggregate(steps ~ interval, activity, FUN = "mean")
print(head(intervalSteps))
print(tail(intervalSteps))
# create a time series plot 
plot(intervalSteps$interval, intervalSteps$steps, type='l',
     main="Average Number of Steps over All Days", xlab="Interval", ylab="Average Number of Steps")

```

```{r}

# Let us first find the row with max number of steps
maxRowSteps <- which.max(intervalSteps$steps)

# Now find the interval with above max
intervalSteps[maxRowSteps, ]

```

## Imputing missing values

We next take the issue of imputing missing values, as there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data. We do the following:

1.	Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

2.	Devise a strategy for filling in all of the missing values in the dataset. WE use the mean for that 5-minute interval.

3.	Create a new dataset that is equal to the original dataset but with the missing data filled in.

4.	Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

We do notice that these values differ significantly in the interval from 10k to 15k where the frequency has increased from 23 to 35 as compared to the estimates from the first part of the assignment.

We notice that due to imputing missing data,  mean values stays the same but therer is slight difference in meadian value.

```{r}
# As we have already ignored the missing values. So, let us again read the csv # file, this time keeping the missing values:
activity <- read.csv("activity.csv")
sum(is.na(activity))

activity2 <- activity
for (i in 1:nrow(activity2)) {
    if (!is.na(activity2[i,"steps"])) next
    intervalValue <- activity2[i,"interval"]
    stepsValue <- intervalSteps[which(intervalSteps$interval == intervalValue), 2]
    #cat(intervalValue,stepsValue,"\n")
    activity2[i,"steps"] <- stepsValue
}

#head(activity2)
#tail(activity2)
sum(is.na(activity2))

# preparing data for time series plot

eachDay2 <- group_by(activity2, date)
stepsTakenEachDay2 <- summarise(eachDay2, total = sum(steps))
head(eachDay2)
tail(eachDay2)
head(stepsTakenEachDay2)
tail(stepsTakenEachDay2)

hist(stepsTakenEachDay2$total, main="Histogram of the total number of steps taken each day", 
     xlab="Total number of steps taken each day")

cat("Mean value of total number of steps each day is", mean(stepsTakenEachDay2$total),"\n")
cat("Median value of total number of steps each day is", median(stepsTakenEachDay2$total))

```

## Are there differences in activity patterns between weekdays and weekends?

We next consider the following question:

Are there differences in activity patterns between weekdays and weekends?

For this part instead of using weekdays() function, we decided to use wday() function. We also Use the dataset with the filled-in missing values for this part.

1. Here we first create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

2.	We then, make a panel plot containing a time series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
library(lubridate)
library(ggplot2)
activity2['dayType'] <- wday(as.Date(activity2$date))
activity2$dayType[activity2$dayType %in% c(1,7) ] <- "weekend"
activity2$dayType[activity2$dayType != "weekend"] <- "weekday"

# convert dayType from cha to factor
activity2$dayType <- as.factor(activity2$dayType)

# calculate average steps by interval across all days
activity2IntervalBySteps <- aggregate(steps ~ interval + dayType, activity2, "mean")

# creat a plot
qplot(interval, 
      steps, 
      data = activity2IntervalBySteps, 
      type = 'l', 
      geom=c("line"),
      xlab = "Interval", 
      ylab = "Number of steps", 
      main = "") +
  facet_wrap(~ dayType, ncol = 1)
```
# From the above plot it is clear that during weekend the data is more spread as compared to weekdays. 
